\textbf{\sc Towards Deliberative Control of Autonomous Robots}

### 2012-02-20 trunk/arch.tex
typos

\item \emph{Partitioning} the global planning problem into multiple
  decision loops, called {\em reactors}. Each reactor has its own
  scope both functionally (one for example focusing on the conversion
  of high level waypoints into low-level commands) and temporally as

  each reactor declares its planning look-ahead $\pi$ and expected
  latency $\lambda$
  % \comment{$\pi \& \lambda$ both defined here for the first time}
  before being able to produce plan to the agent depending on where it
  is located in a reactor dependency graph such as shown in
  Fig. \ref{fig:agent}.

\item \emph{Coupling} a tighter integration between the planning
  process and world evolution to ensure that planning occurs while
  being continuously informed of state change. By doing so, the
  planner is better informed in its deliberation despite a larger
  latency than changing world change state.

@@

The \rx architecture is structured around the notion of \eu {\em state
  variables} as the fundamental basis of interaction. Additionally it
comes with a well-defined ownership model associated with each state
variable. In this framework each reactor is seen as a 'black box'
providing its own {\em internal} state variables which it is
responsible for in order to maintain its value at every instant of
time. Each such instance is represented by a {\em tick} with a fixed
duration; in our implementation of \rx on the Dorado, we use a tick of
$1$ second. In order to identify its state a reactor can subscribe to
{\em external} state variables owned by other reactors, for which it
will receive new updates when the owner reactor generates its {\em
  internal} state.

This {\em synchronization} process propagates the current state
information at the execution frontier $\tau$ throughout all the 
reactors. The sequencing of the reactors synchronization is identified
by traversing the reactor graph from the least dependents reactors --
the ones with no {\em external} timelines -- to the most dependent
ones which typicallu do not provide {\em internal} timelines.
Conversely each reactor can request a future change on one of 
its {\em external} state variables which will be transfromed as 
an {\em internal} goal to the owner. 
Such \emph{goal dispatching} is expected to produce a
deliberation phase for the receiving reactor which can cyclically and
in turn, produce a plan impacting its {\em external} state variables' 
future that will propagate down following the same mechanism.

@@

\item \texttt{synchronize} takes the last {\em external} observations
  as an argument and returns the {\em internal} observations for a
  reactor or a failure report.

\item \texttt{step} that accepts new goals requested (if any) on the
  {\em internal} timeline of a reactor and will execute one step of
  deliberation. In the context of \eu for example one deliberation
  step corresponds to a single flaw resolution (Section
  \ref{sec:europa:arch}) while ensuring that the resulting plan is not
  proven inconsistent. This implies that it also includes the
  backtracking of the search in such a case.  The returned value by
  the \texttt{step} function is a boolean which indicates whether the
  reactor needs an extra step to deliberate or if this last step
  resulted in a complete plan within the reactor's look-ahead. On the
  later, the part of the plan that concern this reactor {\em external}
  timelines is provided in order to be dispatched as goals to
  whichever reactor that is reponsible for these. \comment{By
    extending I may have saud things I developed below ... will have
    to check}
\end{itemize}

Both these deliberative functions have different scope and temporal
constraints. The \texttt{synchronize} call is considered as atomic and
will focus only on state inference for the current tick by identifying
the current value of {\em internal} state variables from the latest
updates on the {\em external} state variables. This process propagates
up in the reactor dependency graph at every tick to ensure that all
reactors are aware of the current state of the world. \texttt{step}
relates to a single decision step of the planning process which is
focused on the future evolution of state variables; it can take
multiple steps spanning multiple ticks for a reactor to produce a plan
that is both complete and valid. When such plan is identified
\comment{how? Is there a terminating condition? No more flaws but I
  washoping it would have been defined in \eu section} its {\em external}
timeline(s) are then dispatched to the corresponding reactors which
own these times and which in turn can start their own deliberation
\texttt{step}s to satisfy these new objectives. Consequently execution
occurs as these goals and plans propagate down in the reactor
hierarchy until they eventually become goals for the lowest level
reactor which -- instead of deliberating -- may just transform this
goal into a command for execution in hardware.

@@

While the two processes are separate, they share the same plan
internal to a reactor; consequently at every synchronization cycle the
planning process is informed of the new world state and its impact on
the plan. Conversely, when synchronization occurs after deliberation
steps, its starting point is the last partial plan resulting from
deliberation. This gives an initial frame to the problemm that helps
-- at least initial -- to target the synchronization problem by
ensuring that the new {\em external} state updates are compatible with
this plan. If so, the synchronization propagates these events in the
plan and can uses it as a help on identifying its {\em internal}
timelines based on this plan. If they are not compatibles, then
synchronization identifies a plan inconsistency  reflecting that the
current plan is not valid anymore. \comment{fpy: I am here and will
  develop further}

\comment{following is not what I intended to express whatsoever
  ... synchroniuzation od not deal with dispatching but execution
  tracking ... it is though helped by the structure of the exisiting
  plan and often identifies execution failures}
 Conversely, when the deliberation process eventually finds a
solution, synchronization is informed about generating the planned
state values \comment{should this read as ``...is informed about
  generated state values..'' instead?} on the {\em external} state
variables. This {\em external} plan defines the set of {\em goals} for
this state variable managed by another {\em reactor}.

Synchronization offers an important challenge for embedding planning
in a situated agent, typically %  as always been to go around the asumptions made by
% most classical planners. Specifically one
an assumption to restrict the problem to ``offline planning'' defined
in \cite{ghallab04} as follows:

\begin{quotation}
  The planner is not concerned with any change that may occur in
  $\Sigma$ \footnote{The world modeled by the plan domain.} while it is
  planning; it plans for the given initial and goal states regardless
  of the current dynamics, if any. 
\end{quotation}

This assumption, while helpful to reduce the scope of the planning
problem, becomes problematic for planning within a highly dynamic
environment. In a typical robot this implies that the only point where
the planner can be informed about the current world state is prior to
the commencement of a planning cycle by giving to the planner the
initial state and current objectives. During the planning, it is
assumed that the agent can maintin the world as perceived by the
planner as stable. \cite{lemai04, lemai-chenevier2004} attempts to
reduce the impact of planning by allowing local plan repair in the
existing plan. In the eventuality that such repair is not feasible
they require that their robot comes to a full stop until (re-)planning
is complete. \comment{I may need more examples/refs here}. As Fig.
\ref{fig:tick-exec} shows, this assumption is problematic for our
architecture. Synchronization that tracks state evolution can occur
several times during the reactor deliberation process. By making
synchronization and deliberation share the same plan we provide a
solution that relaxes this static world assumption. \comment{Need to
  talk about IDEA and the fact that they had similar aspect even
  though they were never very clear nor well articulated in their
  papers}

In the following sections we show how synchronization and deliberation
are implemented independently to show how their interaction through
the plan helps relax this assumption by making planning fully
integrated in the reactor execution cycle while potentially allowing
for more informed planning.

\subsubsection{Synchronization identify internal state evolution}
\label{sec:arch:synch}

We first show how a reactor can track and identify its state in the
context where it does not have any compelling need to deliberate. We
assume that we have a reactor that has no future goal. Even in this
condition, \rx requires each reactor to produce at every tick its {\em
  internal} state. For those reactors not owning {\em internal}
timelines, \rx makes sure that {\em external} observations are taken
into account for every tick as shown in \cite{py10}. Such a
requirement is enforced to ensure that all reactors share a consistent
view of the world with advancing time.

{\em External} state updates are managed by introducing these
observation in the plan structure as facts and ensure that they remain
compatible with the current plan mainainted by the reactor. The
challenging part of synchronization is to decide the {\em internal}
state values for the current tick. In the Shopping example for example
we have the following code snippet which helps identify when we can
\texttt{Buy} a product:

@@

If \texttt{object.location} is {\em external} and the \texttt{Agent}
is {\em internal}, a robot may encounter the situation where it is at
the location where it can buy one (possibly among others)
product. This results om a decision problem related to the state value
of the \texttt{Agent} timeline where the robot can either {\em Buy} a
specific product now, wait a little longer before doing so or just
simply not buy any product at this location.

In the \eu based reactor this decision requirement is a new type of
flaw which forces full identification of internal state of the reactor
for the current tick. Using this new flaw we can describe the
synchronization process as:

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
  \caption{\small Illustration of the synchronization flaws in a reactor. The
    reactor receive new observations when they are produced by the
    owner(s) of its internal timelines. The line after the last token
    of each timeline represent the domain of possible values for the
    end of this token. At every tick $\tau$ the reactor needs to
    integrate the {\em External} state information it received and --
    based on its model $\mathcal{M}$ -- resolve its {\em Internal}
    state that will then be provided by the architecture to other
    reactors using these state variables.}
  \label{fig:synch:flaw}
\end{figure}

We introduce this requirement as a new type of flaw for the \eu
framework defined as follow:

\begin{definition}
  \label{def:csf}
  A {\em current state flaw} indicates that a reactor {\em internal}
  timeline does not provide a fully defined state value for the
  execution frontier $\tau$. Such a flaw can be result of either of
  the following reasons:

  \begin{itemize}
  \item The absence of a token in this timeline which overlaps the
    instant $\tau$.
  \item An open decision between terminating the last token for this
    timeline or pushing its end in the future.
  \end{itemize}

The core idea between this flaw is to ensure that at any single tick
the {\em synchronization} solver will identify all the {\em internal}
timelines for which the current state value is yet to be
identified. We handle their resolution based on the following
proposition.

\begin{proposition}
  \label{prop:csf:resolve}
  A {\em current state flaw} can be resolved using one of the
  following choices which are evaluated in the following order:

@@

  \item Extend the previous state value to end after the execution 
    frontier  (\ie restrict its end time to $[\tau+1, \infty)$). %\comment{first use of
    % $\tau$, so needs to be defined earlier}

  \item Create and insert a new token in this timeline that will start
    at the current tick $\tau$ (attempt this for each possible token
    type for this timeline if necessary).

\end{enumerate}
\end{proposition}
Intuitively the first choice enforces an intertial value assumption
where the reactor will tend to maintain its current as long as the
model and external observations allow it. The second choice favorise
the advance of the reactor current plan if any; this direct the {\em
  synchronization} decision toward assuming that its current plan is
properly executed. Finally if none of these two choices did provide a
valid solution -- meaning that both lead to an inconsistant plan --
the reactor tries all the possible state values remaining for this
timeline until it finds one that allows it to identify its state as
required by \rx architecture.

This flaw forces the reactor to identify its internal state
for the current tick. By using this flaw one can describe the
synchronization process as:

\begin{enumerate}
\item Integrate the external state as provided by the owner of each
  external timeline into the plan database.
\item Propagate this information in the plan database following the
  model $\mathcal{M}$ of this reactor.
\item Resolve the current state value of each internal timelines based
  on Proposition \ref{prop:csf:resolve}.
\end{enumerate}

Steps $2$ \& $3$ are handled by the \eu solver from algorithm
\ref{alg:europa:solve}. We only enforce that the {\em current state
  flaws} are selected last by the algorithm in order to follow the
sequencing given above. As {\em synchronization} focusses on the
execution frontier wee also filter out all flaws that are outside of
the current tick scope. This mean that if this flaw relates to a token
that is {\em necessarily} ending before $\tau$ or starting after
$\tau+1$ then the solver will ignore it. This helps to reduce the
scope of the synchronization to exclusively the current state of the
reactor and reduce dramatically the search space for the {\em
  synchronization} solver. In order to do so we need to assume 
that the current state value of an {\em internal} timeline does not 
depend on the future or more accurately that any choices made during 
this {\em synchronization} will not lead to a future inconsistency. 
Such an assumption impacts the set of possible domains the reactor 
can support while remaining complete. Take for  example this rule from 
our \texttt{Shopping} model :

While this model is perfectly acceptable for planning it becomes
problematic when execution is taken into account especially since it
puts a strong emphasis between an evolving plan and an unknown future
(\texttt{Going} to a location and the {\em expected} future outcome of
ending at this location, ties it strongly to the constraint in line
$5$ above).  So for example, assume that the \texttt{Agent.location}
is {\em external} to the reactor which receives the observation
\texttt{Going(Home,SuperMarket)} which synchronization has resolved by
producing the observation \texttt{Go(Home, SuperMarket)} in the
\texttt{Agent} timeline. This would imply that the only possible
outcome on \texttt{Agent.location} is to be \texttt{At(SuperMarket)}
in the directly foreseeable future. In the real world the fact that we
are \texttt{Going} to a certain location does not necessarly imply
that we will succeed on ending up there. In one extreme case due to an
exogenous event, the road to the \texttt{SuperMarket} is closed which
could result in an alternate observation that the robot is
\texttt{At(7th st. \& Main)}. This observation contradicts the
existing plan the reactor is maintaining. Further, it is important to
understand the temporal direction of the \emph{causal link} that is
triggering the inconsistency. At the time the robot observes this
contradicting the \texttt{At} predicate, the rule that generated this
conflict is connected to past observations \texttt{Going} and
\texttt{Go} which cannot be easily retracted without a potential
global impact to the plan implying a huge cost in back-propagation
that would be problematic.

>>>>

\item \emph{Partitioning} the global planning problem into multiple
  decision loops, called {\em reactors}. Each reactor has its own
  scope both functionally (one for example focusing on the conversion
  of high level waypoints into low-level commands) and temporally as
  each reactor declares its planning look-ahead $\pi$ and expected
  latency $\lambda$ before being able to produce plan to the agent
  depending on where it is located in a reactor dependency graph such
  as shown in Fig. \ref{fig:agent}. A reactor closer to the hardware
  in the hierarchy, is expected to deliberate less and be highly
  responsive to observations from the world. This implies lower values
  of $\pi$ and $\lambda$ (see Section \ref{sec:arch:trex} below).

\item \emph{Coupling} a tighter integration between the planning
  process and world evolution to ensure that planning occurs
  \emph{during continuous} state update. By doing so, the planner is
  better informed in its deliberation despite a larger latency than
  the changing world.

@@

The \rx architecture is structured around the notion of \eu {\em state
  variables} as the fundamental basis of interaction. Additionally it
comes with a well-defined ownership model associated with each such
state variable. In this framework each reactor is seen as a 'black
box' providing its own {\em internal} state variable which it is
responsible for in order to maintain its value at every instant of
time. Each such instance is represented by a {\em tick} with a fixed
duration; in our implementation of \rx on the Dorado AUV, we use a
tick of $1$ second. In order to identify its state a reactor can
subscribe to {\em external} state variables owned by other reactors,
for which it will receive new updates when the owner reactor generates
its {\em internal} state.

\begin{figure}[!htbp]
  \centering
  \includegraphics[scale=0.35]{figs/DAG-hierarchy}
  \caption{\small An example of a \rx reactor hierarchy which is used
    to determine a synchronization order for propagation of values in
    \emph{external} state variables.}
  \label{fig:reactor-DAG}
\end{figure}

As the execution frontier, $\tau$ advances, the \rx agent propagates
the current state information across all reactor state variables
through the {\em synchronization} process.  The sequencing of the
reactors synchronization is identified by traversing a reactor
dependency graph from the least dependent -- ones with no {\em
  external} timelines -- to the most, \ie which typically do not have
{\em internal} timelines. Fig. \ref{fig:reactor-DAG} shows an example
of such a dependency graph implemented as a directed acyclic graph.

Conversely each reactor can request a future change on one of its {\em
  external} state variables which will be transformed as an {\em
  internal} goal to the owner reactor(s).  Such \emph{goal
  dispatching} is expected to produce a deliberation phase for the
receiving reactor which can cyclically in turn, produce a plan
impacting its {\em external} state variables' future that will
propagate down following the same mechanism.

In such a process, one challenge is to avoid being overwhelmed by
goals that may threaten the ability of a reactor to process and
integrate in its evolving partial plan. For that reason, each reactor
specifies the following two parameters that applied towards the scope
of the agent lifetime:

\item $\lambda$ is the specified latency of this reactor, which is an
  indicator of the \emph{maximum expected time} for a reactor to
  produce a plan.\footnote{Note that while it is highly recommended to
    select a sound value, a failure to produce a plan within a chosen
    value for this parameter is not considered as a critical failure
    of the reactor.}

\item $\pi$ is the planning \emph{look-ahead} of the reactor. which is
  indicative of the number of ticks in the future this reactor is
  looking ahead during deliberation.

Using this information along with the dependency relation between
reactors, \rx defines the notion of execution latency of a state
variable as follow:

\begin{definition}
  The {\em execution latency} of a {\em state variable} $s$ which is
  internal to the reactor $r$ is the maximum expected delay before a
  goal on this state variable can be included in its plan and executed
  by the reactor $r$. It is defined recursively as: 
  \begin{equation*}
    \lambda_{exec}(s) = \lambda(r) + \max_{e \in exts(r)}
    \lambda_{exec}(e) 
  \end{equation*}
  Where  $exts(r)$ is the set of the state variables which are
  {\em externals} to $r$. 
\end{definition}

For intertwined plan execution, the concept of execution latency is
critical at the agent level since it can impact the
planning/deliberation process on {\em external} state variables of a
reactor. For simplicity we abstract it out as being equivalent to the
reactor latency implying that we assume that all {\em external}
timelines of this reactor have $\lambda_{exec} = 0$. At the \rx agent
level this execution latency is used to define the {\em dispatch
  window} of the same state variable.% : \comment{Needed to say that
  % early on but the flow of this paragraph is not very tight}
%\kcomment{Looks fine to me unless you want to clarify further.}

\begin{definition}
  \label{def:dispatch}
  For every timeline $s$ {\em internal} to a reactor $r$, the {\em
    dispatch window} of this timeline at the execution frontier $\tau$
  is the time interval:
  \begin{equation*}
    \Pi(s) = [\tau + \lambda_{exec}(s), \tau+\lambda_{exec}(s)+\pi(r)]
  \end{equation*}
\end{definition}

This window is used by \rx to identify when to dispatch a request to a
{\em reactor} as a new goal. This happens only if the request start
time -- represented as the interval of possible start times --
intersects this window. Requests further in the future are buffered by
the agent until this time window advances for an overlap. This
provides an elegant mechanism to request a change from the {\em
  external} state variable of a reactor while avoiding flooding the
owner reactor of this state variable.

The choice of both the latency and look-ahead of a reactor can be
delicate as one is dependent on the other. Needless to say, latency is
highly dependent on exogenous paramaters such as the host computers
computing power and the number of reactors needing to deliberate at
once, among others. As the design stands, these variables should be
evaluated carefully by the system designer balancing agent efficacy.

% efficacy of the agent to find and execute a
% global strategy they are just indicative and a failure from a reactor
% to produce a plan in its latency is not considered as a critical
% failure from \rx point of view.

Requests from a reactor can be {\em recalled} when it is determined
that they are not part of its projected plan anymore. For example if a
reactor needs to replan due to an unexpected event, it will recall all
requests it produced that have not been executed yet.
% \kcomment{replace
%   ``observed'' with ``not incorporated by the owner reactor''?}
% yet
 This recall is sent to the reactor that received the goal(s) to
remove it/them from its current set of objectives.

@@

\item \texttt{synchronize} takes the last set of {\em external}
  observations as an argument and returns the {\em internal}
  observations for a reactor or a failure report.

\item \texttt{step} that accepts new goals requested (if any) on the
  {\em internal} timeline of a reactor and will execute one step of
  deliberation. In the context of \eu for example, one deliberation
  step corresponds to a single flaw resolution (Section
  \ref{sec:europa:arch}) while ensuring that the resulting plan is not
  inconsistent. Backtracking is included in the step should it be such
  a case.  The returned value by the \texttt{step} function is a
  boolean which indicates whether the reactor needs an extra step to
  deliberate or if this last step resulted in a complete plan within
  the reactor's look-ahead. The part of the plan that covers the
  reactor's {\em external} timelines is provided to be dispatched as a
  goal.
% to whichever reactor that is responsible for these. % \comment{By
    % extending I may have saud things I developed below ... will have
    % to check}

Both these 
% deliberative \kcomment{``deliberative??''; something else
%   or remove altogether perhaps?} 
functions have different scope and temporal constraints. 
The \texttt{synchronize} call is considered as atomic and will 
focus only on state inference for the current tick by
identifying the current value of {\em internal} state variables from
the latest updates on {\em external} state variables. This process
propagates up in the reactor dependency graph at every tick to ensure
that all reactors are aware of the current state of the
world. \texttt{step} relates to a single decision step of the planning
process focused on the future evolution of state variables; it can
take multiple steps spanning multiple ticks for a reactor to produce a
plan that is both complete and valid. How the reactor identifies that
its plan is complete and valid is its responsibility and depends on
the way each reactor resolves its plan at each {\em step}. The general
idea is that the reactor is able to find a valid strategy in order to
satisfy as many of its {\em internal} goals as possible. For \eu, this
means that the solver is unable to identify any further flaws to be
resolved in the plan.

When such plan is identified, its {\em external} timeline(s) are then
dispatched to the corresponding reactors which own these times and
which in turn can start their own deliberation \texttt{step}s to
satisfy these new objectives. Consequently execution occurs as these
goals and plans propagate down in the reactor hierarchy until they
eventually become goals for the lowest level reactor which -- instead
of deliberating -- may just transform this goal into a command for
execution in hardware.

@@

While the two processes are separate, they share the same plan
internal to a reactor; consequently at every synchronization cycle the
planning process is informed of the new world state and its impact on
the plan. Conversely, when synchronization occurs after deliberation
stops, its starting point is the last partial plan resulting from
deliberation. This gives an initial frame to the problem that helps to
target the synchronization problem by ensuring that the new {\em
  external} state updates are compatible with this plan. If so,
synchronization propagates these events in the plan and can uses it as
an aid in resolving its {\em current state flaw} by using the plan
information on the exepected {\em internal} state evolution nearby the
execution frontier. 
% \kcomment{``use it
%   as an aid to identify'' what though?}

If they are not compatible, then {\em synchronization} identifies a
plan inconsistency. This results in the retraction of decisions made
during previous {\em deliberation} steps. % in order to
% allow {\em synchronization} to identify its {\em internal} state.
Such a relaxation can be viewed as a backtracking % on the plan search
% triggered exogenously by the world evolution. It is the way our
allowing the reactor to identify that its current plan has failed to
execute triggering further deliberations steps to find an alternate
strategy.

Finally, when the deliberation process has identified a complete
solution for its current {\em internal} goals, a part of its plan that
applies to its {\em external} timelines is sent as a request to \rx to
handle their dispatch to corresponding reactors.  Synchronization
eventually receives {\em external } observations that confirms plan
execution, in turn recording execution progress by having the part of
the plan observed restricted to observation start times.

Synchronization offers an important challenge for embedding planning
in a situated agent, typically an assumption to restrict the problem
to ``offline planning'' defined in \cite{ghallab04}:

{\scriptsize
  \begin{quote}
    The planner is not concerned with any change that may occur in
    $\Sigma$ \footnote{The world modeled by the plan domain.} while it
    is planning; it plans for the given initial and goal states
    regardless of the current dynamics, if any.
\end{quote}}

This assumption, while helpful to reduce the scope and complexity of
the planning problem, becomes problematic for planning within a highly
dynamic environment. In a typical robot this implies that the only
point where the planner can be informed about the current world state
is prior to the commencement of a planning cycle, by providing the
planner an initial state and current objectives. During the planning,
it is assumed that the agent can maintain the world as perceived by
the planner. \cite{lemai04, lemai-chenevier2004} attempts to reduce
the impact of planning by allowing local plan repair in the existing
plan. In the eventuality that such repair is not feasible they require
that their robot comes to a stop until (re-)planning is complete.

This assumption is problematic for situated agents in general and as
Fig. \ref{fig:tick-exec} shows, for \rx. Synchronization that tracks
state evolution can occur several times during reactor
deliberation. By making synchronization and deliberation share the
same plan we provide a solution that relaxes this static world
assumption. \texttt{IDEA} uses the same principle with all solvers in
an agent sharing the same plan for communicating environmental change
\cite{Dias:2003ua, mus06}. Specifically their {\em Reactive planner}
-- similar to our {\em synchronization} solver -- was often the solver
targeting the change at the execution frontier. 
% Although, none of the papers we found were detailing how this was
% handled in the architecture.
\cite{Dias:2003ua} states:

{\scriptsize
  \begin{quote}
  [\textellipsis] as long as the planning horizon over which the
  deliberative planner is working never intersects the current
  execution time, deliberative planning can operate in parallel with
  reactive execution and does not affect the reactivity of the agent. 
\end{quote}}

This suggests a deliberate non overlap between {\em reactive planner}
and {\em deliberative planners} horizons. In \rx, and more
specifically in our \eu reactor the planning horizon of {\em
  synchronization} -- limited to the current tick -- is a subset of
the planning horizon of the deliberation steps -- starting at the
current tick and with an extent based on latency $\lambda$ and
look-ahead $\pi$ of the reactor. In this manner we avoid creating a
conflict between the deliberation and synchronization by controlling
the scheduling of these two process by the \rx agent.

In the following sections we show how synchronization and deliberation
are implemented independently to show how their interaction through
the plan helps relax this assumption. It does so by integrating
planning in the reactor execution cycle even while allowing for more
informed deliberation.

\subsubsection{Synchronization and Identifying Internal State}
\label{sec:arch:synch}

We first show how a reactor can track and identify its state in the
context where it does not have any compelling need to deliberate. In
doing so we assume that we have a reactor that has no future
goal. Under such a condition, \rx requires each reactor to produce at
every tick its {\em internal} state. For those reactors not owning
{\em internal} timelines, \rx makes sure that {\em external}
observations are taken into account for every tick as shown in
\cite{py10}. Such a requirement is enforced to ensure that all
reactors share a consistent view of the world with advancing time.

{\em External} state updates are managed by introducing these
observations in the plan structure as facts and ensuring that they
remain compatible with the current plan maintained by the reactor. The
challenging part of synchronization is to decide the {\em internal}
state values for the current tick. In the Shopping example for example,
we have the following code snippet which helps identify when we can
\texttt{Buy} a product:

@@

If \texttt{object.location} is {\em external} and \texttt{Agent} is an
{\em internal} timeline, a robot may encounter the situation where it
is at the location where it can buy one (possibly among others)
product. This results in a decision-making problem related to the
state value of the \texttt{Agent} timeline; the robot can either {\em
  Buy} a specific product now, wait a little longer before doing so or
just simply not buy any product at this location.

In \texttt{T-REX}'s \eu reactor this decision-making requirement is a
new type of flaw which requires to identify the value of all the {\em
  internal} state variables for the current tick.
% forces identification \kcomment{identification
%   of what? Probably need a little more explanation here.} of internal
% state of the reactor for the current tick. 
This flaw can be defined as:

\begin{definition}
  \label{def:csf}
  A {\em current state flaw} indicates that a reactor's {\em internal}
  timeline does not provide a fully defined state value for the
  execution frontier $\tau$. Such a flaw can be the result of either:

  \begin{itemize}
  \item The absence of a token in this timeline which overlaps the
    execution frontier $\tau$.
  \item An open decision between terminating the last token for this
    timeline or extending its end in the future.
  \end{itemize}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.5\columnwidth]{figs/synch-relation}
  \caption{\small Illustration of synchronization flaws in a
    reactor. The reactor receives new observations when they are
    produced by the owner(s) of its internal timelines. The line
    \kcomment{what line???} after the last token of each timeline
    represent the domain of possible values for the end of this
    token. At every tick $\tau$ the reactor needs to integrate the
    {\em External} state information it received and -- based on its
    model $\mathcal{M}$ -- resolve its {\em Internal} state that will
    then be provided by the agent to other reactors using these state
    variables.}
  \label{fig:synch:flaw}
\end{figure}

The key idea is to ensure that at any single tick the {\em
  synchronization} solver will identify all the {\em internal}
timelines for which the current state value is yet to be
identified. We handle their resolution based on the following
proposition:

\begin{proposition}
  \label{prop:csf:resolve}
  A {\em current state flaw} can be resolved using one of the
  following choices which are evaluated in order:

  \item Extend the previous state value to end after the execution
    frontier (\ie restrict its end time to $[\tau+1, \infty)$).

  \item Create and insert a new token in this timeline that will start
    at the current tick $\tau$ and attempt this for each possible
    token type for this timeline if necessary.

\end{enumerate}
\end{proposition}

Intuitively the first choice enforces an \emph{inertial value
  assumption} where the reactor will tend to maintain its current
state as long as the model and external observations allow for it. The
second choice favors the advance of the reactor's current plan if any;
this directs the {\em synchronization} decision toward assuming that
its current plan is properly executed. Finally if none of these two
choices provides a valid solution -- \ie lead to an inconsistant state
-- the reactor tries all the possible state values remaining for the
timeline until it finds one that brings plan consistency.

% allows it to identify its state as required by \rx architecture.

% The flaw forces the reactor to identify its internal state for the
% current tick. By using this flaw one can describe the synchronization
% process as:

The synchronization process in turn can be described as:

@@

% \begin{enumerate}
% \item Integrate the external state as provided by the owner of each
%   external timeline into the plan database.
% \item Propagate this information in the plan database following the
%   model $\mathcal{M}$ of this reactor.
% \end{enumerate}

Steps $2$ \& $3$ above are handled by the \eu solver from algorithm
\ref{alg:europa:solve} which ensures that the {\em current state
  flaws} are selected last so as to follow the sequence given
above.

As {\em synchronization} focuses on the execution frontier it also
filters out all flaws that are outside of the current tick's scope.
If the flaw correlates to a token that is {\em necessarily} ending
before $\tau$ or starting after $\tau+1$ then the solver ignores
it. Such a strategy helps to reduce the scope of synchronization
exclusively to the current state of the reactor and in turn reduces
the search space for the {\em synchronization} solver.  To do so we
need to assume that the current state value of an {\em internal}
timeline does not depend on the future or more accurately that any
choice made during this {\em synchronization} will not lead to a
future inconsistency.  Such an assumption impacts the set of possible
domains the reactor can support while remaining complete. Take for
example this rule from our \texttt{Shopping} model:

While this model is perfectly acceptable for planning it becomes
problematic when execution is taken into account especially since it
puts a strong emphasis between an evolving plan and an unknown future;
\texttt{Going} to a location and the {\em expected} future outcome of
ending at this location, ties it strongly to the constraint in line
$5$ above. For instance, assume that the \texttt{Agent.location} is
{\em external} to the reactor which receives the observation
\texttt{Going(Home,SuperMarket)} which synchronization has resolved by
producing the observation \texttt{Go(Home, SuperMarket)} in the
\texttt{Agent} timeline. This would imply that the only possible
outcome on the \texttt{Agent.location} timeline, is to be
\texttt{At(SuperMarket)} in the directly foreseeable future.

In the real world the fact that the robot is \texttt{Going} to a
certain location does not necessarly imply that it will succeed in
ending up there. Due to an exogenous event, for example, the road to
the \texttt{SuperMarket} could be closed, which could result in an
alternate observation that the robot is \texttt{At(7th st. \& Main)}
contradicting the existing plan. Further, it is important to
understand the temporal direction of the \emph{causal link} that is
triggering the inconsistency. At the time the robot observes the
contradiction of the \texttt{At} predicate, the rule that generated
this conflict is connected to past observations \texttt{Going} and
\texttt{Go}. These cannot be retracted easily without a potential
impact to the plan implying in terms of a cost in back-propagation.

More generally tying a token to future outcomes in a model that is
meant to be confronted by execution is problematic. It might be valid
if one can guarantee that an outcome is predisposed to occur (for
example the reactor managing the \texttt{Agent.location} timeline
guarantees that the outcome of the \texttt{Going} will {\em always}
result on ending at the traget location). However, given the very
nature of the planning problem and the ability to search outcomes and
to replan, the model should be written so that a future projection can
be manipulated.

One could therefore alter this part of the model to give the reactor
more flexibility as follows:

We do so by relaxing the constraint in line $5$ from a strict
\texttt{equals} to a \texttt{contains} and to add the new constraint
at line $11$ which ensures that the last \texttt{Going} ends up in the
location the agent actually wants the robot to be at. This first level
of relaxation will therefore allow the agent to use multiple
\texttt{Going} tokens should the first one fail to reach the
objective.

As the simple example above shows, synchronization involves a
deliberation process with the risk that the search for a solution is
not a straightforward sequence of staright-forward decisions where 
the resolution can be easily identified by the solver as any other
option would leafd immediately to an inonsistency in the plan.
% \kcomment{has a unit decision been explained in Europa?}. 
Since such a process is required by the architecture at \emph{every}
tick for \emph{every} reactor the model needs to be designed carefully.

One critical aspect to take into account is that from a \rx point of
view, synchronization is the only critical task of a reactor. Such a
process is used in order for the agent to identify its current state
and maintain it consistently throughout the agent's lifetime. For a
reactor not being able to find a consistent solution during
synchronization -- implying that after evaluating all possible
solutions provided by a model it was unable to find a consistent plan
-- not only indicates that a reactor is unable to explain the state of
the world, but more importantly it may jeopardize other reactors that
rely on its {\em internal} state. For this reason a failure to
synchronize immediately results in the \rx agent killing a reactor and
notifying any reactor that depends on its {\em internal} state
variables; it does so by posting the special observation
\textsf{Failed}. By so doing we allow for graceful degradation where
other reactors can readapt their plans on receipt of this observation.

Finally, at the engine level careful choices were made in both the
focus of the search -- limited only to tokens that can potentially
overlap the current tick -- and heuristics where the {\em current
  state flaw} is considered last during the search process. The later
limits the amount of backtracking for a solution -- or at least how
deep one needs to backtrack in a decision tree.  Similarly the
sequence in which we evaluate the possible resolution for such a flaw
-- presented in Proposition \ref{prop:csf:resolve} -- was selected
according to some simple assumptions, that any external observation
that is compatible with an existing token in the current plan reflects
the execution of the plan itself. In other words this means that our
solver always prefers to use tokens \emph{already existing} in the
plan as a way to identify current internal state. By doing so
synchronization exploits the structure of the plan produced during
previous deliberation steps.

% This implies that a deliberation steps
% have been done and did impact the plan structure and calls for a
% description on how our \eu reactor resolve such steps which we are now
% going to describe further.

Deliberation inside a reactor is based on \eu capabilities described
in Section \ref{sec:europa:arch}. However this deliberation comes with
some augmentation namely the reactor's deliberation latency and its
plan look-ahead both expressed in ticks and that the agent needs to
interleave deliberation with not just the synchronization but also
with deliberation of other reactors\footnote{The current
  implementation of \rx is running on a single process and it is the
  responsibility of the agent itself to emulate reactor
  multi-threading for deliberation and synchronization.}.  For this
reason while the reactor relies on the \eu solver, it slices the
execution of Algorithm \ref{alg:europa:solve} into atomic steps. This
interruption occurs around the recursive call line \ref{li:recurse}
while ensuring that the current partial plan is not inconsistant at
any single step. This is managed by allowing the call to backtrack
until a \textsf{DecisionPoInt} that is not exhausted (\ie
$decision.hasNext()$ is \texttt{true}) is found in the call stack.

A critical aspect of this deliberation step is to ensure that at the
exit of this call the $plan$ is not inconsistent. % Indeed, it is always
% possible that the next call made by the agent is for synchronization
% which could fail if the $plan$ is inconsistent before any attempt is
% made for synchronization.
If the partial plan is found to be inconsistent during a deliberation
step we either backtrack in the decision stack until a node of the
tree is found with an alternate solution or the plan is relaxed since
there is no other alternative solution. We do this by removing all
goals while keeping only recent observations.

{\em Synchronization} and the dispatching of new {\em goals} from \rx
can occur between two steps and alter the plan. These alterations
exogenous to the deliberation process dynamically impact the set of
flaws that need to be resolved. New flaws are created when a new goal
is received from \rx and conversely the recall of an objective could
remove the corresponding flaws\footnote{However if the flaws related
  to this goal were resolved, the removal of this goal may likely
  create new flaws.}. The impact of {\em synchronization} can be more
subtle as its alteration of the plan at the execution frontier is more
drastic. We can still give a non exhaustive list of possible
alterations of the flaws directly related to {\em synchronization}:

@@

\item Previous flaws directly related to the execution frontier have
  been resolved by the {\em synchronization} solver.

\item The {\em synchronization} solver may have created new flaws that
  were outside of its scope but are now part of the scope of the {\em
    deliberation} solver.

\item The last observation has proved that the current plan is invalid
  requiring the {\em synchronization} solver to relax the existing
  plan to identify its current {\em external} state. (This case is
  developed further on the next section).
